FastSRM: A fast, memory efficient and identifiable implementation of the shared response model

### 2022-07-28 paper/main.tex
edited the intro

In this work, we implement an identifiable version of SRM and show on real data
that it improves the stability of the recovered shared response.
%
We then introduce FastSRM, which relies on a dimension reduction step that we
prove to yield the same solution as the original algorithm.
% be optimal in the sense that working with reduced data does not induce any change in the
% algorithm trajectory.
%
We show experimentally using synthetic and real fMRI data
that FastSRM is considerably faster and more memory efficient
than current implementations.

@@

\section{Introduction}
%\bt{We need to start with a more application-oriented intro}
%
Complex stimulations such as movie-watching, story or music listening are more
and more popular in the neuro-scientific community. Indeed such naturalistic
paradigms are unconstrained from behavioral manipulations and thus, more
ecological with respect to every-day cognitive conditions.
%
Yet, in contrast to standard designs, these so-called
naturalistic stimuli do not come with a design matrix that quantifies the
features of interest in the stimuli.
%
Without a design matrix, models such as the general linear
model~\cite{poline2012general} cannot be used.
%
Although some works have used manual
annotations~\cite{huth2012continuous} or deep neural
networks~\cite{gucclu2017increasingly}~\cite{richard2018optimizing},
see also \url{https://neuroscout.org}, to create an estimate of the
design matrix of naturalistic stimuli, it is a hard task.

Another approach is to learn jointly the design matrix and the spatial maps in
an unsupervised way.
%
In the case where the dataset contains only one subject,
independent component analysis~\cite{jutten1991blind} (ICA) is the method of choice.
%
ICA assumes that components are independent, a defendible assumption
which ensures identifiability up to permutation and scaling of the
model and can be fitted efficiently.
%
When the dataset contains multiple subjects undergoing the same
protocol, a natural assumption is to assume that the design matrix is
shared across subjects.
%
Many different methods can produce shared components from different
subjects.
%
Some assume independent
components~\cite{richard2021model}~\cite{richard2020modeling}~\cite{varoquaux2009canica}~\cite{calhoun2001method},
while others are only based on second order information such as
Multiset Canonical Correlation Analysis~\cite{via2011joint} or the
Shared Response Model (SRM)~\cite{chen2015reduced}.
%

>>>>

In this work, we implement an identifiable version of SRM and show on real data
that it improves the stability of the recovered shared response.
%
We then introduce FastSRM, which relies on a dimension reduction step that we
prove to yield the same solution as the original algorithm.
%
We show experimentally using synthetic and real fMRI data
that FastSRM is considerably faster and more memory efficient
than current implementations.

@@

\section{Introduction}
%\bt{We need to start with a more application-oriented intro}
%
Complex stimulations such as movie-watching, story or music listening are more
and more popular in the neuro-scientific community. Indeed such naturalistic
paradigms are unconstrained from behavioral manipulations and thus, more
ecological with respect to every-day cognitive conditions.
%
If one wants to use data acquired under naturalistic conditions to map
functional responses to the brain, i.e. in an encoding setting
\cite{naselaris2011}, one has to deal with the fact that no design matrix is
specified for naturalistic stimuli.
%
Such a design matrix encodes the temporal events that affect brain
signal during an experiment; these events typically reflect the occurrence
of some features of interest in the stimuli.
%
Although some works have used manual
annotations~\cite{huth2012continuous} or deep neural
networks~\cite{gucclu2017increasingly}~\cite{richard2018optimizing},
see also \url{https://neuroscout.org}, to create an estimate of the
design matrix of naturalistic stimuli, it is a hard task.
%
Without a design matrix, models such as the general linear
model~\cite{poline2012general} cannot be used.

Another approach is to learn jointly a set of reference time courses  and the spatial maps in
an unsupervised way.
%
These reference time courses formally replace the design matrix in the
spatio-temporal decomposition of the data, but cannot be considered as
a formal design matrix.
%
In the case where the dataset contains only one subject,
independent component analysis~\cite{jutten1991blind} (ICA) is the method of choice.
%
ICA assumes that components are independent, a defendible assumption
which ensures identifiability up to permutation and scaling of the
model and can be fitted efficiently.
%
When the dataset contains multiple subjects undergoing the same
protocol, a natural assumption is to assume that the temporal reference is
shared across subjects.
%
Many different methods can produce shared components from different
subjects.
%
Some assume independent
components~\cite{richard2021model}~\cite{richard2020modeling}~\cite{varoquaux2009canica}~\cite{calhoun2001method},
while others are only based on second order information such as
Multiset Canonical Correlation Analysis~\cite{via2011joint} or the
Shared Response Model (SRM)~\cite{chen2015reduced}.
%
